---
title: "CellMentor Test"
format: html
editor: visual
---

## Running CellMentor Code


```{r}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("petrenkokate/CellMentor", dependencies = TRUE)
```

```{r}
install.packages("rlang")
```

## Basic Usage

Set random seed for reproducibility

```{r}
set.seed(100)
```

Load required packages

```{r}
library(CellMentor)
library(Matrix)
library(ggplot2)
library(dplyr)
library(Seurat)
library(scRNAseq)
```

### 1. Load the Datasets
```{r}
# Loading reference dataset (Baron)
baron <- hBaronDataset()
reference_matrix <- baron$data
reference_celltypes <- baron$celltypes

# Loading query dataset (Muraro)
muraro <- muraro_dataset()
query_matrix <- muraro$data
query_celltypes <- muraro$celltypes # This would be unknown in a real application
                                            # We keep it here for evaluation
```

### (Optional) Create smaller subsets for faster demonstration
```{r}
# Function to create balanced subsets
create_subset <- function(matrix, celltypes, cells_per_type = 30) {
  # Get unique cell types
  unique_types <- unique(celltypes)
  
  # Select cells for each type
  selected_cells <- c()
  for (cell_type in unique_types) {
    # Get cells of this type
    type_cells <- names(celltypes)[celltypes == cell_type]
    
    # If fewer cells than requested, take all of them
    n_to_select <- min(cells_per_type, length(type_cells))
    
    # Randomly select cells
    selected <- sample(type_cells, n_to_select)
    selected_cells <- c(selected_cells, selected)
  }
  
  # Return subset
  list(
    matrix = matrix[, selected_cells],
    celltypes = celltypes[selected_cells]
  )
}

# Create balanced subsets with 30 cells per type
baron_subset <- create_subset(reference_matrix, reference_celltypes, 30)
muraro_subset <- create_subset(query_matrix, query_celltypes, 30)

# Update variable names for clarity
reference_matrix <- baron_subset$matrix
reference_celltypes <- baron_subset$celltypes
query_matrix <- muraro_subset$matrix
query_celltypes <- muraro_subset$celltypes # This would be unknown in a real application
                                            # We keep it here for evaluation
```


### 2. Create a CellMentor Object
```{r}
csfnmf_obj <- CreateCSFNMFobject(
  ref_matrix = reference_matrix,
  ref_celltype = reference_celltypes,
  data_matrix = query_matrix,
  norm = TRUE,
  most.variable = TRUE,
  scale = TRUE,
  scale_by = "cells",
  verbose = TRUE,
  num_cores = 1
)
```

### 3. Run CellMentor and choose optimal params

The parameter selection process involves optimizing multiple hyperparameters to achieve the best cell type separation. Note that the function is computationally intensive and because of it we test only limited ranges of parameters.

Parameter descriptions:
- alpha_range: Controls within-class scatter (cell similarity within the same type)
- beta_range: Controls between-class scatter (cell separation between different types)
- gamma_range: Controls sparsity of the factorization
- delta_range: Controls orthogonality between factors

These are parameter settings that work well across most datasets without extensive tuning.

```{r}
# Find optimal parameters
optimal_params <- CellMentor(
  csfnmf_obj,
  alpha_range = c(1, 5),      # Limited alpha range
  beta_range = c(1, 5),       # Limited beta range
  gamma_range = c(0.1),     # use only one gamma for speed
  delta_range = c(1),       # use only one delta for speed
  num_cores = 5,
  verbose = TRUE
)

# Get best model
best_model <- optimal_params$best_model
K_VALUE <- best_model@parameters$rank
```


Check best params
```{r}
print(optimal_params$best_params)
```

### 4. Project Data

```{r}
# Project query data onto the learned space
h_project <- project_data(
  W = best_model@W, # Learned gene weights
  X = best_model@matrices@data,  # Query data matrix
  num_cores = 5,
  verbose = TRUE
)
```

### 5. Integration with Seurat
```{r}
rownames(query_matrix) <- make.unique(rownames(query_matrix))
seu_muraro <- CreateSeuratObject(counts = query_matrix)
seu_muraro$celltype <- query_celltypes
# Add CellMentor dimensionality reduction to Seurat object
seu_muraro$CellMentor <- CreateDimReducObject(
    embeddings = t(as.matrix(h_project)),
    key = "CellMentor_",
    assay = DefaultAssay(seu_muraro),
    loadings = as.matrix(best_model@W)
)

# Visualization
seu_muraro <- RunUMAP(seu_muraro, reduction = 'CellMentor', dims= 1:K_VALUE)
DimPlot(seu_muraro, group.by = 'celltype')
```
