---
title: "CellMentor - Testing with Pulmonary Ionocyte Dataset"
format: html
editor: visual
---

# CellMentor - Testing with Pulmonary Ionocyte Dataset

## Install / load packages

```{r}
required_pkgs <- c("Matrix", "ggplot2", "dplyr", "Seurat", "CellMentor")
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "CellMentor") {
      if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
      BiocManager::install("CellMentor")
    } else {
      install.packages(pkg)
    }
  }
}

library(Matrix)
library(ggplot2)
library(dplyr)
library(Seurat)
library(CellMentor)

set.seed(100)
```

## Source the extension

```{r}
source("extension/diagnostics.R")
source("extension/plotting.R")
```

## Download the data

```{r}
DATA_URL <- paste0(
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE103nnn/",
  "GSE103354/suppl/GSE103354_Trachea_droplet_UMIcounts.txt.gz"
)
local_file <- "GSE103354_Trachea_droplet_UMIcounts.txt.gz"

if (!file.exists(local_file)) {
  message("Downloading GSE103354 droplet UMI counts …")
  download.file(DATA_URL, destfile = local_file, mode = "wb")
}
```

## Load and inspect the count matrix

```{r}
message("Reading count matrix …")
raw <- read.table(
  gzfile(local_file),
  header      = TRUE,
  sep         = "\t",
  row.names   = 1,
  check.names = FALSE
)
message(sprintf("Matrix dimensions: %d genes × %d cells", nrow(raw), ncol(raw)))
```

## Parse cell-type labels from column names

```{r}
col_names <- colnames(raw)
celltypes <- sub(".*_", "", col_names)   # last field  → cell type
mouse_ids <- sub("_.*", "", col_names)   # first field → M1–M6 (batch)

type_counts <- sort(table(celltypes), decreasing = TRUE)
message("Cell type counts:")
print(type_counts)

message("Mouse ID / batch counts:")
print(table(mouse_ids))
```

## Convert to sparse matrix

```{r}
counts <- Matrix(as.matrix(raw), sparse = TRUE)
```

## Reference / Query split

```{r}
create_split <- function(counts, celltypes, mouse_ids, ref_frac = 0.70, seed = 100) {
  set.seed(seed)
  all_types <- unique(celltypes)
  ref_idx   <- c()
  qry_idx   <- c()
  for (ct in all_types) {
    idx    <- which(celltypes == ct)
    n_ref  <- max(1, round(length(idx) * ref_frac))
    chosen <- sample(idx, n_ref)
    ref_idx <- c(ref_idx, chosen)
    qry_idx <- c(qry_idx, setdiff(idx, chosen))
  }
  list(
    ref_counts    = counts[, ref_idx],
    ref_celltypes = setNames(celltypes[ref_idx], colnames(counts)[ref_idx]),
    ref_mouse_ids = mouse_ids[ref_idx],
    qry_counts    = counts[, qry_idx],
    qry_celltypes = setNames(celltypes[qry_idx], colnames(counts)[qry_idx]),
    qry_mouse_ids = mouse_ids[qry_idx]
  )
}

splits <- create_split(counts, celltypes, mouse_ids, ref_frac = 0.70)

ref_counts    <- splits$ref_counts
ref_celltypes <- splits$ref_celltypes
qry_counts    <- splits$qry_counts
qry_celltypes <- splits$qry_celltypes

message(sprintf(
  "Reference: %d cells | Query: %d cells",
  ncol(ref_counts), ncol(qry_counts)
))
message("Cell-type distribution in reference:")
print(table(ref_celltypes))
message("Cell-type distribution in query:")
print(table(qry_celltypes))
```

## Create the CSFNMF object

```{r}
message("Creating CSFNMF object …")
csfnmf_obj <- CreateCSFNMFobject(
  ref_matrix    = ref_counts,
  ref_celltype  = ref_celltypes,
  data_matrix   = qry_counts,
  norm          = TRUE,
  most.variable = TRUE,
  scale         = TRUE,
  scale_by      = "cells",
  verbose       = TRUE,
  num_cores     = 10
)
```

## Run CellMentor - parameter search

```{r}
message("Running CellMentor parameter search …")
optimal_params <- CellMentor(
  csfnmf_obj,
  alpha_range = c(1, 5),
  beta_range  = c(3, 4),
  gamma_range = c(0.1),
  delta_range = c(1),
  num_cores   = 10,
  verbose     = TRUE
)

best_model <- optimal_params$best_model
K_VALUE    <- best_model@parameters$rank

message("Best parameters:")
print(optimal_params$best_params)
message(sprintf("Optimal rank K = %d", K_VALUE))
```

## Project query data onto learned latent space

```{r}
message("Projecting query data …")
h_project <- project_data(
  W         = best_model@W,
  X         = best_model@matrices@data,
  num_cores = 10,
  verbose   = TRUE
)
```

## Extract matrices for the extension

```{r}
# These are the preprocessed (normalised, HVG-filtered, scaled) matrices
# stored inside the fitted model
X_query <- best_model@matrices@data   # genes × n_query (preprocessed query)
X_ref   <- best_model@matrices@ref    # genes × n_ref   (preprocessed reference)
H_ref   <- best_model@H               # K × n_ref       (reference embeddings)

# Reference cell type labels aligned to columns of X_ref
ref_ct <- tryCatch(
  best_model@annotation$celltype,
  error = function(e) ref_celltypes[colnames(X_ref)]
)
```

## Run projection diagnostics

```{r}
diag_out <- run_projection_diagnostics(
  W                 = best_model@W,
  H_ref             = H_ref,
  X_ref             = X_ref,
  ref_celltypes     = ref_ct,
  H_query           = h_project,
  X_query           = X_query,
  novelty_threshold = 0.95,       # flag cells above 95th percentile of reference error
  novelty_method    = "per_class",
  distance_metric   = "cosine",
  verbose           = TRUE
)

diagnostics <- diag_out$diagnostics
```

## Attach true labels and save

```{r}
# Attach ground-truth labels for evaluation
diagnostics$true_label <- qry_celltypes[diagnostics$cell]

write.csv(diagnostics, "ionocyte_diagnostics.csv", row.names = FALSE)

message("Diagnostics saved to ionocyte_diagnostics.csv")
message(sprintf(
  "%d / %d query cells flagged as potential novel (%.1f%%)",
  sum(diagnostics$is_potential_novel),
  nrow(diagnostics),
  100 * mean(diagnostics$is_potential_novel)
))
```

## Novelty summary by cell type

```{r}
diagnostics |>
  group_by(true_label) |>
  summarise(
    n          = n(),
    n_novel    = sum(is_potential_novel),
    pct_novel  = round(100 * mean(is_potential_novel), 1),
    mean_error = round(mean(reconstruction_error), 4),
    .groups    = "drop"
  ) |>
  arrange(desc(pct_novel)) |>
  print()
```

## Extension plots

```{r}
# Error distributions: reference vs. query with novelty threshold marked
p_err <- plot_error_distributions(
  ref_errors   = diag_out$ref_errors,
  query_errors = diagnostics$reconstruction_error,
  threshold    = quantile(diag_out$ref_errors, 0.95)
)
ggsave("ionocyte_error_distributions.png", p_err, width = 6, height = 4, dpi = 150)
p_err
```

```{r}
# Confidence histogram colored by novelty flag
p_conf <- plot_confidence_distribution(diagnostics)
ggsave("ionocyte_confidence_distribution.png", p_conf, width = 6, height = 4, dpi = 150)
p_conf
```

## Seurat integration & UMAP visualisation

```{r}
message("Building Seurat object and running UMAP")

rownames(qry_counts) <- make.unique(rownames(qry_counts))
seu <- CreateSeuratObject(counts = qry_counts)
seu$celltype <- qry_celltypes
seu$mouse_id <- splits$qry_mouse_ids

seu[["CellMentor"]] <- CreateDimReducObject(
  embeddings = t(as.matrix(h_project)),
  key        = "CellMentor_",
  assay      = DefaultAssay(seu),
  loadings   = as.matrix(best_model@W)
)

seu <- RunUMAP(seu, reduction = "CellMentor", dims = 1:K_VALUE)
```

## UMAP diagnostic overlays

```{r}
# Align diagnostics rows to the UMAP cell order
umap_coords         <- Seurat::Embeddings(seu, "umap")
diagnostics_ordered <- diagnostics[match(rownames(umap_coords), diagnostics$cell), ]

panels <- plot_umap_diagnostics(
  umap_coords = umap_coords,
  diagnostics = diagnostics_ordered,
  true_labels = diagnostics_ordered$true_label
)

# Save and display each panel
ggsave("ionocyte_umap_true.png",      panels$true,       width = 6, height = 5, dpi = 150)
ggsave("ionocyte_umap_predicted.png", panels$predicted,  width = 6, height = 5, dpi = 150)
ggsave("ionocyte_umap_confidence.png",panels$confidence, width = 6, height = 5, dpi = 150)
ggsave("ionocyte_umap_error.png",     panels$error,      width = 6, height = 5, dpi = 150)
ggsave("ionocyte_umap_novelty.png",   panels$novel,      width = 6, height = 5, dpi = 150)

panels$true
panels$predicted
panels$confidence
panels$error
panels$novel
```

## Standard cell type UMAP and ionocyte highlight

```{r}
p_ct <- DimPlot(seu, group.by = "celltype", label = TRUE, repel = TRUE) +
  ggtitle("CellMentor UMAP – Mouse Tracheal Epithelium (Query)\nColored by cell type") +
  theme(legend.position = "right")
p_ct
```

```{r}
seu$is_ionocyte <- seu$celltype == "Ionocyte"
p_ion <- DimPlot(
  seu,
  group.by = "is_ionocyte",
  cols     = c("TRUE" = "#E63946", "FALSE" = "grey80"),
  order    = "TRUE",
  pt.size  = 1.2
) +
  ggtitle("Ionocytes highlighted on CellMentor UMAP") +
  theme_bw()
ggsave("ionocytes_highlighted.png", p_ion, width = 6, height = 5, dpi = 150)
p_ion
```

## Quantitative Evaluation (ARI and NMI)

```{r}
if (!requireNamespace("aricode", quietly = TRUE)) install.packages("aricode")
library(aricode)

evaluate_clustering <- function(seu, K, resolution = 0.1) {
  seu <- FindNeighbors(seu, reduction = "CellMentor", dims = 1:K, verbose = FALSE)
  seu <- FindClusters(seu, resolution = resolution, verbose = FALSE)

  clusters   <- as.integer(seu$seurat_clusters)
  truth      <- as.integer(factor(seu$celltype))
  ari        <- ARI(clusters, truth)
  nmi        <- NMI(clusters, truth)
  n_clusters <- length(unique(clusters))
  n_types    <- length(unique(seu$celltype))

  message(sprintf(
    "resolution=%.2f | clusters=%d (true types=%d) | ARI=%.3f | NMI=%.3f",
    resolution, n_clusters, n_types, ari, nmi
  ))
  list(seu = seu, ari = ari, nmi = nmi,
       n_clusters = n_clusters, resolution = resolution)
}
```

```{r}
message("\nClustering evaluation at resolution = 0.1")
eval_res <- evaluate_clustering(seu, K_VALUE, resolution = 0.1)
```

```{r}
message("\nResolution sweep:")
resolutions   <- c(0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5)
sweep_results <- lapply(resolutions, function(r) evaluate_clustering(seu, K_VALUE, resolution = r))

sweep_df <- data.frame(
  resolution = sapply(sweep_results, `[[`, "resolution"),
  n_clusters = sapply(sweep_results, `[[`, "n_clusters"),
  ARI        = sapply(sweep_results, `[[`, "ari"),
  NMI        = sapply(sweep_results, `[[`, "nmi")
)
message("\nResolution sweep summary:")
print(sweep_df)
```

```{r}
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
library(tidyr)

sweep_long <- pivot_longer(sweep_df, cols = c("ARI", "NMI"),
                           names_to = "metric", values_to = "score")

p_sweep <- ggplot(sweep_long, aes(x = resolution, y = score,
                                  colour = metric, group = metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.11, y = 0.1, label = "paper default",
           hjust = 0, colour = "grey40", size = 3.5) +
  scale_colour_manual(values = c("ARI" = "#2166ac", "NMI" = "#d6604d")) +
  labs(x = "Clustering resolution", y = "Score", colour = "Metric",
       title = "CellMentor clustering performance vs resolution") +
  theme_bw(base_size = 12)
ggsave("performance_vs_resolution.png", p_sweep, width = 6, height = 5, dpi = 150)
p_sweep
```

```{r}
summary_df <- data.frame(
  resolution = eval_res$resolution,
  ARI        = round(eval_res$ari, 3),
  NMI        = round(eval_res$nmi, 3),
  n_clusters = eval_res$n_clusters
)
message("\nFinal clustering metrics")
print(summary_df)
```
