---
title: "CellMentor - Testing with Pulmonary Ionocyte Dataset"
format: html
editor: visual
---

# CellMentor - Testing with Pulmonary Ionocyte Dataset

## Install / load packages

```{r}
required_pkgs <- c("Matrix", "ggplot2", "dplyr", "Seurat", "CellMentor")
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "CellMentor") {
      if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
      BiocManager::install("CellMentor")
    } else {
      install.packages(pkg)
    }
  }
}

library(Matrix)
library(ggplot2)
library(dplyr)
library(Seurat)
library(CellMentor)

set.seed(100)
```

## Download the data

The droplet UMI count matrix is \~13 MB gzipped - manageable. Rows = genes, Columns = cells (tab-separated, first row = header)

```{r}
DATA_URL <- paste0(
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE103nnn/",
  "GSE103354/suppl/GSE103354_Trachea_droplet_UMIcounts.txt.gz"
)
local_file <- "GSE103354_Trachea_droplet_UMIcounts.txt.gz"

if (!file.exists(local_file)) {
  message("Downloading GSE103354 droplet UMI counts …")
  download.file(DATA_URL, destfile = local_file, mode = "wb")
}
```

## Load and inspect the count matrix

```{r}
message("Reading count matrix …")
raw <- read.table(
  gzfile(local_file),
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)
# Dimensions: ~18,000 genes × ~7,193 cells
message(sprintf("Matrix dimensions: %d genes × %d cells", nrow(raw), ncol(raw)))
```

```{r}
raw
```

## Parse cell-type labels from column names

```{r}
# Column names have the format:  <MouseID>_<Barcode>_<CellType>
# e.g. "M1_GCTTGAGAAGTCGT_Club"  or  "M3_TTTGCGCGTTAAGAAA_Ionocyte"
# M1–M6 are individual mouse IDs; the cell type is the last field.
col_names   <- colnames(raw)
celltypes   <- sub(".*_", "", col_names)        # last field  → cell type
mouse_ids   <- sub("_.*", "", col_names)        # first field → M1 … M6 (batch)
 
# How many of each type?
type_counts <- sort(table(celltypes), decreasing = TRUE)
message("Cell type counts:")
print(type_counts)
# Expected: Basal, Club, Ciliated, Goblet, Tuft, Neuroendocrine,
#           Ionocyte, Hillock-associated, Squamous, etc.
 
message("Mouse ID / batch counts:")
print(table(mouse_ids))
```

## Convert to sparse matrix

```{r}
counts <- Matrix(as.matrix(raw), sparse = TRUE)
```

## Reference / Query split

```{r}
# Strategy: use 70 % of each cell type as REFERENCE, 30 % as QUERY.
# This ensures every cell type (including rare ionocytes) appears in both.

create_split <- function(counts, celltypes, mouse_ids, ref_frac = 0.70, seed = 100) {
  set.seed(seed)
  all_types <- unique(celltypes)
  ref_idx   <- c()
  qry_idx   <- c()
  for (ct in all_types) {
    idx     <- which(celltypes == ct)
    n_ref   <- max(1, round(length(idx) * ref_frac))
    chosen  <- sample(idx, n_ref)
    ref_idx <- c(ref_idx, chosen)
    qry_idx <- c(qry_idx, setdiff(idx, chosen))
  }
  list(
    ref_counts    = counts[, ref_idx],
    ref_celltypes = setNames(celltypes[ref_idx], colnames(counts)[ref_idx]),
    ref_mouse_ids = mouse_ids[ref_idx],
    qry_counts    = counts[, qry_idx],
    qry_celltypes = setNames(celltypes[qry_idx], colnames(counts)[qry_idx]),
    qry_mouse_ids = mouse_ids[qry_idx]
  )
}
 
splits <- create_split(counts, celltypes, mouse_ids, ref_frac = 0.70)
 
ref_counts    <- splits$ref_counts
ref_celltypes <- splits$ref_celltypes
qry_counts    <- splits$qry_counts
qry_celltypes <- splits$qry_celltypes
 
message(sprintf(
  "Reference: %d cells | Query: %d cells",
  ncol(ref_counts), ncol(qry_counts)
))
message("Cell-type distribution in reference:")
print(table(ref_celltypes))
message("Cell-type distribution in query:")
print(table(qry_celltypes))
```

## Create the CSFNMF object

```{r}
message("Creating CSFNMF object …")
csfnmf_obj <- CreateCSFNMFobject(
  ref_matrix    = ref_counts,
  ref_celltype  = ref_celltypes,
  data_matrix   = qry_counts,
  norm          = TRUE,     # library-size normalization
  most.variable = TRUE,     # select highly variable genes
  scale         = TRUE,
  scale_by      = "cells",
  verbose       = TRUE,
  num_cores     = 10         # increase/decrease depending on your # of cores
)
```

## Run CellMentor - parameter search

```{r}
# Using the "safe" default ranges recommended in the paper's supplement.
# Reduce ranges further (e.g. alpha_range = c(1), beta_range = c(1)) if slow.
message("Running CellMentor parameter search …")
optimal_params <- CellMentor(
  csfnmf_obj,
  alpha_range = c(1, 5),
  beta_range  = c(6, 7),
  gamma_range = c(0.1),
  delta_range = c(1),
  num_cores   = 10,
  verbose     = TRUE
)
 
best_model <- optimal_params$best_model
K_VALUE    <- best_model@parameters$rank
 
message("Best parameters:")
print(optimal_params$best_params)
message(sprintf("Optimal rank K = %d", K_VALUE))
```

## Project query data onto learned latent space

```{r}
message("Projecting query data …")
h_project <- project_data(
  W        = best_model@W,
  X        = best_model@matrices@data,
  num_cores = 10,
  verbose  = TRUE
)
```

## Seurat integration & UMAP visualization

```{r}
message("Building Seurat object and running UMAP")

rownames(qry_counts) <- make.unique(rownames(qry_counts))
seu <- CreateSeuratObject(counts = qry_counts)
seu$celltype  <- qry_celltypes
seu$mouse_id  <- splits$qry_mouse_ids

# Add CellMentor reduction
seu[["CellMentor"]] <- CreateDimReducObject(
  embeddings = t(as.matrix(h_project)),
  key        = "CellMentor_",
  assay      = DefaultAssay(seu),
  loadings   = as.matrix(best_model@W)
)

# UMAP on CellMentor dimensions
seu <- RunUMAP(seu, reduction = "CellMentor", dims = 1:K_VALUE)
```

## Plots

```{r}
# UMAP colored by annotated cell type
p1 <- DimPlot(seu, group.by = "celltype", label = TRUE, repel = TRUE) +
  ggtitle("CellMentor UMAP – Mouse Tracheal Epithelium (Query)\nColored by cell type") +
  theme(legend.position = "right")

p1
```

```{r}
# UMAP - highlight ionocytes specifically
seu$is_ionocyte <- seu$celltype == "Ionocyte"
p2 <- DimPlot(
  seu,
  group.by = "is_ionocyte",
  cols     = c("TRUE" = "#E63946", "FALSE" = "grey80"),
  order    = "TRUE",
  pt.size  = 1.2
) +
  ggtitle("Ionocytes highlighted on CellMentor UMAP") +
  theme_bw()

p2
```

## Quantitative Evaluation (ARI and NMI)
```{r}
# Following the paper's Methods:
#   1. Graph-based clustering on the CellMentor latent space
#   2. Compare clusters to ground-truth cell types using ARI and NMI
# Default resolution in the paper: 0.1

if (!requireNamespace("aricode", quietly = TRUE)) install.packages("aricode")
library(aricode)

evaluate_clustering <- function(seu, K, resolution = 0.1) {
  seu <- FindNeighbors(seu, reduction = "CellMentor", dims = 1:K, verbose = FALSE)
  seu <- FindClusters(seu, resolution = resolution, verbose = FALSE)

  clusters <- as.integer(seu$seurat_clusters)
  truth    <- as.integer(factor(seu$celltype))

  ari <- ARI(clusters, truth)
  nmi <- NMI(clusters, truth)

  n_clusters <- length(unique(clusters))
  n_types    <- length(unique(seu$celltype))

  message(sprintf(
    "resolution=%.2f | clusters=%d (true types=%d) | ARI=%.3f | NMI=%.3f",
    resolution, n_clusters, n_types, ari, nmi
  ))

  list(
    seu        = seu,
    ari        = ari,
    nmi        = nmi,
    n_clusters = n_clusters,
    resolution = resolution
  )
}
```

```{r}
message("\nClustering evaluation at resolution = 0.1")

eval_res <- evaluate_clustering(
  seu,
  K_VALUE,
  resolution = 0.1
)
```

```{r}
# Sweep resolutions to assess clustering stability and match expected cell types
message("\nResolution sweep:")

resolutions <- c(0.05, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5)

sweep_results <- lapply(resolutions, function(r) {
  evaluate_clustering(seu, K_VALUE, resolution = r)
})

sweep_df <- data.frame(
  resolution = sapply(sweep_results, `[[`, "resolution"),
  n_clusters = sapply(sweep_results, `[[`, "n_clusters"),
  ARI        = sapply(sweep_results, `[[`, "ari"),
  NMI        = sapply(sweep_results, `[[`, "nmi")
)

message("\nResolution sweep summary:")
print(sweep_df)
```

```{r}
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
library(tidyr)

sweep_long <- pivot_longer(
  sweep_df,
  cols = c("ARI", "NMI"),
  names_to = "metric",
  values_to = "score"
)

p_sweep <- ggplot(sweep_long, aes(x = resolution, y = score,
                                  colour = metric, group = metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.11, y = 0.1, label = "paper default",
           hjust = 0, colour = "grey40", size = 3.5) +
  scale_colour_manual(values = c("ARI" = "#2166ac", "NMI" = "#d6604d")) +
  labs(
    x = "Clustering resolution",
    y = "Score",
    colour = "Metric",
    title = "CellMentor clustering performance vs resolution"
  ) +
  theme_bw(base_size = 12)

p_sweep
```

```{r}
summary_df <- data.frame(
  resolution = eval_res$resolution,
  ARI        = round(eval_res$ari, 3),
  NMI        = round(eval_res$nmi, 3),
  n_clusters = eval_res$n_clusters
)

message("\nFinal clustering metrics")
print(summary_df)
```